---
title: "STA2453 Lab 1"
author: "Yihan Duan"
date: "21/09/2021"
output: pdf_document
---

# Exercise 1

## Data Quality

Load raw ED data.
```{r tidy=TRUE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load the required libraries
library(tidyverse)
library(lubridate)

# Read in the data
ed_data <- read_csv('raw_ed_data.csv')

# Show head
head(ed_data)
```

Clean up 'presenting_complaint' column as shown in class.

```{r tidy=TRUE}
# function to clean presenting complaints text
clean_complaints <- function(x) {
  
  x_clean <- x %>% 
    # remove any leading and trailing spaces
    trimws() %>% 
    # collapse > 1 blank space into 1 blank space
    gsub(' +',' ',.) %>% 
    # set text to lower case
    tolower() 
  
  return(x_clean)
  
}

ed_data <- ed_data %>% 
  mutate(presenting_complaint = clean_complaints(presenting_complaint))

ed_data <- ed_data %>% 
  mutate(presenting_complaint = case_when(
    presenting_complaint == "chest pian" ~ "chest pain",
    presenting_complaint == "burns" ~ "burn",
    presenting_complaint == "traumatic injuries" ~ "traumatic injury",
    presenting_complaint %in% c("unk", "missing") ~ "unknown",
    presenting_complaint == "headach" ~ "headache",
    TRUE ~ presenting_complaint))

ed_data %>% 
  count(presenting_complaint, sort = T) %>% 
  mutate(proportion = round(n/sum(n), 3))
```

As shown in class, we can count all the 'NA's for each column.

```{r tidy=TRUE}
count_NAs <- function(x) {
  
  num_NAs <- sum(is.na(x))
  
  return(num_NAs)
}

# the data
ed_data %>% 
  # becomes the first argument passed to the summarize_all function
  summarize_all(count_NAs) %>%
  glimpse
```

```{R tidy=TRUE}
ed_data %>%
  filter(is.na(ed_start_time) & is.na(ed_end_time))
```

Now we check all the variables one by one.

### ENCOUNTER_NUM

First see if there are any duplicate numbers.

```{R tidy=TRUE}
print("Length of ENCOUNTER_NUM:")
length(ed_data$ENCOUNTER_NUM)
print("Number of unique values of ENCOUNTER_NUM:")
length(unique(ed_data$ENCOUNTER_NUM))
```

We can see that there are duplicates in ENCOUNTER_NUM column.

```{R tidy=TRUE}
# Remove all duplicate rows
ed_data_dedup <- ed_data[!duplicated(ed_data),]
```

Now check the number of unique values again.

```{R tidy=TRUE}
print("Length of ENCOUNTER_NUM:")
length(ed_data_dedup$ENCOUNTER_NUM)
print("Number of unique values of ENCOUNTER_NUM:")
length(unique(ed_data_dedup$ENCOUNTER_NUM))
```

Still one duplicate ENCOUNTER_NUM, lets find it.

```{R tidy=TRUE}
num_freq <- ed_data_dedup %>%
  count(ENCOUNTER_NUM) %>%
  filter(n>1)
dup_num = num_freq$ENCOUNTER_NUM[1]
ed_data_dedup %>%
  filter(ENCOUNTER_NUM == dup_num)
```

Notice the only difference between these two records are the `ed_pia_time`.

### CTAS_CD

Check data integrity for column 'CTAS_CD'.

```{R tidy=TRUE}
ed_data_dedup %>% 
  count(CTAS_CD)
```

As we can see, the values presented in this column are mostly in range [1, 5]. However, there are many records with CTAS_CD missing.

### CTAS_DESCR

First check all the values presented in CTAS_DESCR.

```{R tidy=TRUE}
ed_data_dedup %>% 
  count(CTAS_DESCR)
```

There should be a 1 to 1 mapping from CTAS_CD to CTAS_DESCR.

```{R tidy=TRUE}
ed_data[,c("CTAS_CD","CTAS_DESCR")] %>%
  unique
```

Yes the mapping is 1 to 1.

### ed_start_time, ed_end_time, ed_pia_time, adm_start_time

ed_start_time should always come before ed_end_time.

```{R tidy=TRUE}
print("PIA before arrival at the ED:")
ed_data %>%
  filter(ed_start_time > ed_pia_time) %>%
  nrow

print("Departure before arrival at the ED:")
ed_data %>%
  filter(ed_start_time > ed_end_time) %>%
  nrow

print("PIA after leaving the ED:")
ed_data %>%
  filter(ed_pia_time > ed_end_time) %>%
  nrow

print("PIA after admitted to the hospital:")
ed_data %>%
  filter(ed_pia_time > adm_start_time) %>%
  nrow

print("Admitted to the hospital before leaving the ED:")
ed_data %>%
  filter(ed_end_time > adm_start_time) %>%
  nrow

print("Arrived at the ED after admitted to the hospital:")
ed_data %>%
  filter(ed_start_time > adm_start_time) %>%
  nrow

print("PIA time of '2099-01-01':")
ed_data_dedup %>%
  filter(ed_pia_time == ymd("2099-01-01")) %>%
  nrow
```

```{R tidy=TRUE}
ed_data_dedup %>%
  filter(ed_pia_time == ymd("2099-01-01"))
```
As we can see, many 'ed_pia_time' entries are labeled with unrealistic dates -- '2099-01-01'.

### adm_start_time, admitted

adm_start_time only make sense if the patient is admitted.

```{R tidy=TRUE}
ed_data %>%
  filter(admitted == 0) %>%
  count(adm_start_time)
```

All patients that are not admitted have no 'adm_start_tme'.

```{R tidy=TRUE}
ed_data %>%
  filter(admitted == 1) %>%
  count_NAs()
```

There are 371 records that are admitted but have no 'adm_start_time'.

### los
```{R tidy=TRUE}
ed_data %>%
  filter(los == 24) %>%
  count
```

## Descriptive analysis

### Length of stay

```{R tidy=TRUE}
ed_data_dedup$startToEnd = as.numeric(difftime(ed_data_dedup$ed_end_time, ed_data_dedup$ed_start_time), units="hours")

ed_data_dedup %>%
  filter(startToEnd >= 0) %>%
  ggplot(aes(x=startToEnd)) + 
  geom_histogram(binwidth=1) + 
  labs(x = "(Departure Time - Arrival Time) in hours", y="Number of Records")
ggsave("los.png")
```

```{R tidy=TRUE}
ed_data_dedup %>%
  filter(startToEnd >= 0) %>%
  ggplot(aes(x=startToEnd)) + 
  geom_histogram(binwidth=1) + 
  xlim(0, 60) +
  labs(x = "(Departure Time - Arrival Time) in hours", y="Number of Records")
ggsave("los_zoom.png")
```

```{R tidy=TRUE}
startToEndMean = mean(ed_data_dedup$startToEnd[ed_data_dedup$startToEnd > 0], na.rm=TRUE)
startToEndSd = sd(ed_data_dedup$startToEnd[ed_data_dedup$startToEnd > 0], na.rm=TRUE)

startToEndMean
startToEndSd
```

### Time to seeing a physician

```{R tidy=TRUE}
ed_data_dedup$startToPia = as.numeric(difftime(ed_data_dedup$ed_pia_time, ed_data_dedup$ed_start_time), units="mins")

ed_data_dedup %>%
  filter(ed_pia_time != ymd("2099-01-01")) %>%
  filter(startToPia > 0) %>%
  ggplot(aes(x=startToPia)) + 
  geom_histogram(binwidth=10) + 
  labs(x = "(PIA Time - Arrival Time) in mins", y="Number of Records")
ggsave("pia.png")
```

```{R tidy=TRUE}
ed_data_dedup %>%
  filter(ed_pia_time != ymd("2099-01-01")) %>%
  filter(startToPia > 0) %>%
  ggplot(aes(x=startToPia)) + 
  geom_histogram(binwidth=10) + 
  xlim(0,1000) +
  labs(x = "(PIA Time - Arrival Time) in mins", y="Number of Records")
ggsave("pia_zoom.png")
```

```{R tidy=TRUE}
startToPiaMean = mean(ed_data_dedup$startToPia[ed_data_dedup$startToPia > 0 & ed_data_dedup$ed_pia_time != ymd("2099-01-01")], na.rm=TRUE)
startToPiaSd = sd(ed_data_dedup$startToPia[ed_data_dedup$startToPia > 0 & ed_data_dedup$ed_pia_time != ymd("2099-01-01")], na.rm=TRUE)

startToPiaMean
startToPiaSd
```

### Counts and proportions of presenting complaints

```{R tidy=TRUE}
PC_counts = ed_data_dedup %>%
  count(presenting_complaint) %>%
  arrange(desc(n))

jpeg(filename="PC.jpg",width=800,height=600)
pie(PC_counts$n, labels=PC_counts$presenting_complaint, radius = 0.9, cex = 1.2)
dev.off()
```

```{R tidy=TRUE}
PC_counts
```

### CTAS

```{R tidy=TRUE}
CTAS_counts = ed_data_dedup %>%
  count(CTAS_DESCR) %>%
  arrange(desc(n))

jpeg(filename="CTAS.jpg",width=800,height=600)
pie(CTAS_counts$n, labels=CTAS_counts$CTAS_DESCR, radius = 1, cex = 1)
dev.off()
```

```{R tidy=TRUE}
CTAS_counts
```

### Number of encounters

```{R tidy=TRUE}
ed_data_dedup$ed_start_time_YMD = as_date(ed_data_dedup$ed_start_time)

n_encounters = count(ed_data_dedup, ed_start_time_YMD)

ed_data_dedup %>%
  count(ed_start_time_YMD) %>%
  ggplot(aes(x=n)) + 
  geom_histogram(binwidth = 5) +
  labs(x = "Number of Encounters", y="Count")
ggsave("n_enc.png")
```

```{R}
n_encountersMean = mean(n_encounters$n, rm.na=TRUE)
n_encountersSd = sd(n_encounters$n)

n_encounters %>%
  ggplot(aes(x=n)) + 
  geom_density() +
  stat_function(fun=dnorm, args=c(n_encountersMean, n_encountersSd / 1.7), xlim=c(0, 799), n=800, size=1, color='grey') +
  xlim(0,800) +
  labs(x = "Number of Encounters", y="Density", title="Real Density vs. Normal(221, 22)")
ggsave("n_enc_fit.png")
```

```{R}
n_encountersMean
n_encountersSd 
```

### Time series

```{R tidy=TRUE}
ggplot(n_encounters, aes(x=ed_start_time_YMD, y=n)) + geom_line() + labs(x="Date", y="ED arrivals per day", title="ED arrivals per day")
ggsave("ed_arr.png")
```

```{R tidy=TRUE}
n_encountersMean
```

```{R tidy=TRUE}
tail(n_encounters)
```

## Weekend vs. weekday volumes

```{R tidy=TRUE}
weekend_names = c("Sat", "Sun")

# Get weekday and weekend dfs
weekday_encounters = n_encounters %>%
  filter(! weekdays(ed_start_time_YMD, abbreviate = TRUE) %in% weekend_names) %>%
  filter(! is.na(ed_start_time_YMD))
weekend_encounters = n_encounters %>%
  filter(weekdays(ed_start_time_YMD, abbreviate = TRUE) %in% weekend_names)

# Get weekday stats
print("Weekday mean:")
mean(weekday_encounters$n)
print("Weekday standard deviation:")
sd(weekday_encounters$n)

# Get weekend stats
print("Weekend mean:")
mean(weekend_encounters$n)
print("Weekend standard deviation:")
sd(weekend_encounters$n)
```

```{R tidy=TRUE}
weekday_encounters$day = "Weekdays"
weekend_encounters$day = "Weekends"
all_encounters = rbind(weekday_encounters, weekend_encounters)

all_encounters %>%
  ggplot(aes(x=day, y=n)) +
  geom_boxplot() +
  labs(x="Weekday vs. weekend volumes", y="ED arrivals perday")
ggsave("wkd_wke.png")
```

```{R tidy=TRUE}
t.test(n ~ day, data=all_encounters, var.equal=TRUE)
```

## Working hours

```{R tidy=TRUE}
ed_data_clean <- ed_data_dedup %>%
  filter(! is.na(ed_start_time), ! is.na(ed_end_time)) %>%
  filter(ed_start_time < ed_end_time) %>%
  mutate(ed_start_time = floor_date(ed_start_time, unit="hour"),
         ed_end_time = floor_date(ed_end_time, unit="hour"))

arrivals <- ed_data_clean %>%
  select(timestamp=ed_start_time) %>%
  mutate(counter = 1)

departures <- ed_data_clean %>%
  select(timestamp=ed_end_time) %>%
  mutate(counter = -1)

census_volumes <- arrivals %>%
  bind_rows(departures) %>%
  arrange(timestamp, counter) %>%
  mutate(volume = cumsum(counter))

start <- min(ed_data_clean$ed_start_time)
end <- max(ed_data_clean$ed_end_time)
full_time_window <- tibble(timestamp = seq(start, end, by='hours'))

census_volumes <- census_volumes %>%
  right_join(full_time_window,
             by='timestamp') %>%
  arrange(timestamp) %>%
  fill(volume, .direction='down')

census_volumes <- census_volumes %>%
  arrange(timestamp, volume) %>%
  group_by(timestamp) %>%
  summarise_all(last)

census_volumes %>%
  ggplot(aes(timestamp, volume)) +
  geom_line() +
  labs(x='Date', title='Emergency Department Census Throughout 2019')
ggsave("census_ydm_hms.png")
```

```{R tidy=TRUE}
census_hours <- census_volumes %>%
  mutate(timestamp = format(timestamp, format='%T'))

aggregate(x = census_hours$volume, by =list(timestamp=census_hours$timestamp), FUN=mean) %>%
  ggplot(aes(timestamp, x, group=1)) +
  geom_line() +
  labs(y='Volume', x='Time of Day', title='Average ED Census by Hour') +
  theme(axis.text.x = element_text(angle = 45))
ggsave("census_hms.png")
```

```{R}
census_by_hour = aggregate(x = census_hours$volume, by =list(timestamp=census_hours$timestamp), FUN=mean)

summary(census_by_hour)
```

## Descriptive Summary
```{R tidy=TRUE}
ed_data_dedup %>%
  filter(startToEnd >= 0) %>%
  filter(ed_pia_time != ymd("2099-01-01")) %>%
  filter(startToPia >= 0) %>%
  summary()
```






















